Quantitative processing of broadband data as implemented in a scientific splitbeam echosounder

### 2022-04-26 Paper/Manuscript.tex
Edits to sections II, III, and IV

@@

The theoretical foundation for signal processing comes from radar applications and has been documented for broadband echosounders \citep{stanton2008}. To tap the full potential of broadband echosounder data, further data processing methods are being developed \citep{lavery2017,bassett_broadband_2018}. Examples include adapting pulses to improve target separation as well as further parameters from the individual targets. Different transmission pulse forms can be envisioned as well as various methods for acoustic target classification. When translating equations into computer processing code, several choices must be made - some of these are well founded in signal processing literature whereas others are of a more practical and ad-hoc nature. The latter is typically missing in the literature, making it difficult to test the implementation of the signal processing methods in new echo sounders and post processing software.

@@

\subsection{Accompanying code}
The code accompanying this paper is written in Python and is available as supplementary materials. All processing steps with respective figures in the paper can be reproduced by running the main scripts (TSf.py and Svf.py). Without loss of generality, we use the Simrad EK80 as an example since it is currently the most used broadband echosounder in the marine ecosystem acoustics field. By presenting the design goals, implementation details, and recommended procedures and processing required to obtain quantitative broadband data, the authors hope to encourage and facilitate the realistic use of broadband signals in marine ecosystem acoustics.

@@

\subsection{System overview}
A basic echosounder system consists of a transducer, a transceiver, and a computer program that controls the operation of the transceiver and records received signals. During transmission the program defines the signals which are created as electric signals in the transceiver, converted to acoustic signals by the transducer and transmitted into the water. The acoustic signals propagate through the water, are reflected or scattered by objects in the water, and propagate back to the transducer. During reception the transducer converts the received acoustic signals to electric signals, which are received, pre-amplified, filtered, digitized, and processed in the transceiver, and then transferred to the controlling program for further data processing and storage (Fig. \ref{fi:ek_sys}). Many types of transmit signals are feasible - this paper considers only the transmission of linear frequency modulated signals (also known as linear chirps).

@@

The controlling computer program generates a short-duration digital transmit signal (a ping), $\ytx(\samplesymt)$, where $\samplesymt$ is a sample index in the discrete time domain. Typical BB pulses are linear upsweep pulses windowed by an envelope function. The generated signal $\ytxe(\timesym)$ is converted to an analogue electric signal and amplified by the transceiver to obtain analogue signal $\ytxe(\timesym)$. The analogue and amplified signal is passed on to the transducer to generate the transmitted acoustic signal $\ytxa(\timesym)$ in the water column, where $\timesym$ is the time for the (analog) signal. For a split-aperture echosounder system, there are typically three or four channels to allow estimation of the angle of arriving echoes, and the signal is typically transmitted with equal power across the channels.

In the example a linear sweep enveloped by a Hann function is implemented (Fig.~\ref{fi:ytx}), where the parameters are the initial frequency, the final frequency, the pulse duration, the sampling rate, and the proportion of the signal that is tapered in each end, respectively. A tapering of 0 and .5 indicates no tapering and tapering across the whole signal, respectively.
\begin{figure}
\includegraphics[width=16cm]{Fig_ytx}
\caption{\label{fi:ytx} Ideal and enveloped transmit pulse between 34 kHz to 45 kHz with a slope of 0.057, as implemented in the EK80. The orange curve illustrates the effect of setting the slope to 0.5.}
\end{figure}

@@

To remove noise and reduce the quantity of data, the sampled signal from each channel is filtered and decimated in multiple stages, $\stagesym$, using complex bandpass filters, $\hfl(\genidxsym,\stagesym)$, and decimation factors, $\decfac(\stagesym)$. The individual filter coefficients for each filter and decimation stage are indexed by $i$. The output signal from each channel, $\channelsym$, from each filter and decimation stage, $\stagesym$, is then given by:
%
\begin{equation}
\label{eq:yrx}
\yrx(\samplesymt,\channelsym,\stagesym) = \left( \yrx(\samplesymt,\channelsym,\stagesym-1) * \hfl(\genidxsym,\stagesym) \right)_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\yrx(\samplesymt,\channelsym,0)$ is set to $\yrxorg(\samplesymt,\channelsym)$, being the signal before decimation, $*$ indicates convolution and $\nstages$ is the total number of filter stages. The output signal from the final filter and decimation stage, $\yrx(\samplesymt,\channelsym,\nstages)$, is shortened to $\yrx(\samplesymt,\channelsym)$ for convenience. For the output signal, $\yrx(\samplesymt,\channelsym)$, the decimated sampling rate, $\fsdec$, is given by:
%
\begin{equation}
\label{eq:fsdec}
\fsdec = \fs\prod_{\stagesym=1}^{\nstages} \frac{1}{\decfac(\stagesym)}.
\end{equation}
%
The characteristics of the bandpass filter and decimation factors are chosen with regard to the desired operating bandwidth, noise suppression levels, impulse response duration, and other common filter characteristics, with the aim of maintaining sufficient information in the data. In the example $N_v=2$. The frequency responses of the filters are shown in Figure~\ref{fi:fir}, and the corresponding filter coefficients and decimation factors are found in the test data set.

@@

The original sample data $\yrxorg(\samplesymt,\channelsym)$ are not available in the EK80 data files. Instead, the resulting filtered and decimated complex samples from each transducer channel $\yrx(\samplesymt,\channelsym)$ are stored in the data files. The data are recorded together with additional data such as from position and motion sensors and system configuration data in raw data files for display and analysis by processing software. The filter parameters are in this paper used to generate the transmit pulse for the matched filter, see below.

\subsection{Pulse compression}
To increase signal-to-noise ratio and resolution along the acoustic beam a matched filter may be applied to the raw data samples \citep{turin1960}. This technique is also known as pulse compression \citep{klauder1960}. One approach for a matched filter is to use a normalized version of the ideal transmit signal as the replicate signal, filtered and decimated using the same filters and decimation factors as applied in Eq. \ref{eq:yrx}. The normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, is given by:
%
\begin{equation}
\label{eq:ytxnorm}
\ytxnorm(\samplesymt) = \frac{\ytx(\samplesymt)}{\textrm{max}(\ytx(\samplesymt))}\end{equation}
%
where $\textrm{max}$ is the maximum value of $\ytx(\samplesymt)$. The filtered and decimated output signal, $\ytxnorm(\samplesymt,\stagesym)$, from each filter stage, $\stagesym$, using the normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, as the input signal, is given by:
%
\begin{equation}
\label{eq:FilterStagesTX}
\ytxnorm(\samplesymt,\stagesym) = \left[ \ytxnorm(\samplesymt,\stagesym-1) * \hfl(\genidxsym,\stagesym) \right]_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\ytxnorm(\samplesymt,0)$ is set to $\ytxnorm(\samplesymt)$ and $\downarrow$ indicates decimation by the factor $\decfac(\stagesym)$. The output signal from the final filter and decimation stage, $\ytxnorm(\samplesymt,\nstages)$, is used as the matched filter and is denoted as $\ymf(\samplesymt)$  (Fig.~\ref{fi:y_mf_n}).
%
\begin{figure}
\includegraphics[width=16cm]{Fig_y_mf_n}
\caption{\label{fi:y_mf_n} The absolute value of the filtered and decimated output signal $\ymf(\samplesymt)$ used by the pulse compression.}
\end{figure}

@@

To perform pulse compression the received signal, $\yrx(\samplesymt,\channelsym)$, is convolved with a complex conjugated and time-reversed version of the matched filter signal with the matched filter signal, and here also normalized with the $l^2$-norm of the matched filter to maintain received signal power. The pulse compressed signal, $\ypc(\samplesymt,\channelsym)$, then becomes
\begin{equation}
\label{eq:PulseComp}
\ypc(\samplesymt,\channelsym) = \frac{\yrx(\samplesymt,\channelsym)*\ymf^*(-\samplesymt)}{||\ymf||^2_2},
\end{equation}
%
where $||\ymf||$ indicates the $l^2$-norm of $\ymf$, also known as the Euclidean norm. The received power samples are then used to estimate target strength and volume backscattering strength. For estimating received power samples, the mean signal, $\ypc(\samplesymt)$, over all transducer sectors, $\nchannels$, will be used:
%
\begin{equation}
\label{eq:SumSig}
\ypc(\samplesymt) = \frac{1}{\nchannels} \sum_{\channelsym = 1}^{\nchannels} \ypc(\samplesymt,\channelsym).
\end{equation}
%
Compensation of echo strength for position in the acoustic beam requires an estimate of the echo arrival angle. This is obtained using the split-aperture method \citep{burdic1991}, which for broadband pules can be implemented with the angle values contained in the complex-valued $\ypc(\samplesymt)$ data, in combination with knowledge of transducer sector geometry. The principle is demonstrated with a transducer that is divided into four quadrants (Fig. \ref{fi:trd_quad}). In this example the summed signals from four halves (1+2, 2+3, 3+4, 4+1) are calculated as:
\begin{eqnarray}
\label{eq:SumHalves}
y_{\textrm{pc,fore}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,3)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,aft}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,2) \right),\\
y_{\textrm{pc,star}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,port}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,2)+\ypc(\samplesymt,3) \right),
\end{eqnarray}
%
where fore, aft, star(board), and port indicate the relevant transducer halves.
\begin{figure}
\includegraphics{Fig_trd_quad}
\caption{\label{fi:trd_quad}Transducer divided into four quadrants. The labels are directions often used when a transducer is mounted on a ship.}
\end{figure}
%

\subsection{Power and angle samples}
The transceiver measures voltage over a load, $\zrxe$, connected in series with the transducer impedance, $\ztde$. When calculating various acoustic properties a system gain parameter will be used which assumes a matched receiver load. The total received power, $\prxe(\samplesymt)$, from all transducer sectors for a matched receiver load (Fig. \ref{fi:impedances}) is given by: 
\begin{equation}
\label{eq:prx}
\prxe(\samplesymt) = \nchannels\left( \frac{|\ypc(\samplesymt)|}{2\sqrt{2}} \right)^2 \left( \frac{|\zrxe+\ztde|}{\zrxe} \right)^2 \frac{1}{|\ztde|}.
\end{equation}
%
\begin{figure}
\includegraphics[width=16cm]{Fig_impedances}
\caption{\label{fi:impedances}Equivalent circuit diagram of transducer/transceiver with system impedances.}
\end{figure}

Forward/aft and port/starboard phase angles of target echoes are estimated by combining the transducer half signals thus: 
%
\begin{eqnarray}
\label{eq:phase1}
y_\along(\samplesymt) & = & y_{\textrm{pc,fore}}(\samplesymt) y_{\textrm{pc,aft}}^*(\samplesymt), \\
y_\athw(\samplesymt) & = & y_{\textrm{pc,star}}(\samplesymt) y_{\textrm{pc,port}}^*(\samplesymt),
\end{eqnarray}
%
where $y_\along(\samplesymt)$ is the electrical angle along the minor axis of the transducer (positive in the forward direction when ship-mounted) and $y_\athw(\samplesymt)$ the electrical angle along the major axis of the transducer (positive to starboard when ship-mounted), where complex signals are represented in the form $e^{j 2\pi \freqsym \timesym}$, where $j = \sqrt{-1}$. The physical echo arrival angles ($\along$ and $\athw$) are then given by:
%
\begin{eqnarray}
\label{eq:phase2}
\along(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\along(\samplesymt)), \Re(y_\along(\samplesymt) \right)}{\anglefalong}\right) \\
\athw(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\athw(\samplesymt)), \Re(y_\athw(\samplesymt) \right)}{\anglefathw}\right).
\end{eqnarray}
%
where $\anglefalong$ and $\anglefathw$ are constants that convert from phase angles to physical echo arrival angles and are derived from the transducer geometry and $\fc$ the centre frequency of the chirp pulse \citep{ehrenberg1979}. The inverse sine is indicated by $\arcsin$, the four quadrant inverse tangent which returns values in the interval $[-\pi, \pi]$ inclusive is indicated by $\atan$, the real part of a complex number by $\Re$ and the imaginary part by $\Im$. As a mnemonic, the horizontal line in the symbol used for the forward/aft direction, $\along$, represents the pivot axis for the alongship angles and the near-vertical line in the $\athw$ symbol indicates the pivot axis for port/starboard angles.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_theta_phi}
\caption{\label{fi:theta_phi}Physical angles $\along$ and $\athw$.}
\end{figure}

@@

To illustrate each step of the process to obtain $TS(m)$ we use a single ping from a data file where a single metallic calibration sphere with 35 mm diameter (WC35) is suspended approximately 6 m below a 120 kHz transducer (fig.~\ref{Fig_TS_echogram}).

@@

Based on the point scattering strength samples and the phase angle samples, single targets can be detected, and range and bearing to the single targets can be estimated. This is typically achieved through a single echo detection algorithm (SED). Here we will assume that the samples from the pulse compressed data $\ypc(\samplesymt)$ originating from single target already have been identified, noting that the number of samples after the detected target may be higher than those those before the peak to include scattering processes for beyond ideal point targets. The alongship angle $\along_n$, athwartship angle $\athw_n$ and range $r_n$ at the \emph{peak} power $\prxe(\samplesymt)$ within the detected target is used as estimates for $\along_t$, $\athw_t$ and $r_t$, respectively (Fig. \ref{fi:TS}. A simple pseudo SED algorithm is implemented in the code for illustrative purposes.

From the autocorrelation function of the matched filter signal, $\ymfauto(\samplesymt)$, the equivalent number of samples around the peak are extracted to create the reduced autocorrelation signal of the matched filter signal, $\ymfautored(\samplesymt)$. Depending on the target scattering characteristics and the distance to any adjacent single targets, the number of samples around the peak echo level in $\ypctarget(\samplesymt)$ that contain the majority of the echo energy can be more or less than the total number of samples around the peak of $\ymfauto(\samplesymt)$. If the number of samples around the target is more than the total number of samples around the peak of $\ymfauto(\samplesymt)$ all samples around the peak of $\ymfauto(\samplesymt)$ are used. If the number of samples around the target is less than the total number of samples around the peak of $\ymfauto(\samplesymt)$, this lower number is used to create a reduced autocorrelation signal, $\ymfautored(\samplesymt)$.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_singleTarget}
\caption{\label{fi:SED} The $\prxe(\samplesymt)$ power (upper) and split beam angles ($\along_t$ and $\athw_t$) (middle) for the single target. The vertical line corresponds to the range $r_t$ for the single target. The $\ymfautored(\samplesymt)$ (lower) is the autocorrelation function of the transmit signal reduced to the length of the target signal and aligned with the peak power of the target.}
\end{figure}

@@

As a metallic sphere is a rather simple and ideal scatterer, we use $\sv$ from a school of non-swimbladdered fish (fig.~\ref{Fig_Sv_echogram}) collected with a 120 kHz centre frequency transducer to illustrate the processes to obtain $S_v(m)$.

@@

\begin{figure}
\includegraphics[width=16cm]{Fig_Sv_echogram}
\caption{\label{Fig_Sv_echogram} Sv as a function of frequency ($m$) and range ($n$) for a echogram raw data file. The red vertical line indicates ping to be processed in this section.}
\end{figure}

@@

By selecting a set of centre samples $n$, $Sv$ values can be presented as a function of range ($n$) and frequency ($m$) for each ping. The range for centre samples $n$ could be chosen to half the window length or any other grid that the user prefer the data presented to be in. This can be useful when combining the $Sv(m)$ across a range of transducers. In our example we have simply chosen n to be on the same order of the range resolution (fig.~\ref{Fig_Sv_m_n}).

\begin{figure}
\includegraphics[width=16cm]{Fig_Sv_m_n}
\caption{\label{Fig_Sv_m_n} Sv as a function of frequency ($m$) and range ($n$) for a single ping}
\end{figure}

For acoustic abundance estimation and classification purposes it is common to integrate the Sv over a range, in our example we integrate $Sv$ from 15 to 34 m covering a school of non-swimbladdered fish (fig.~\ref{Fig_Sv_echogram}). We note that we normally require averaging over several pings to obtain an unbiased estimate of $Sv$ for echo integration purposes. Here we present the $Sv(m)$ over the layer for our example ping for illustrative purposes (fig.~\ref{Fig_Sv_avg}). Even though this is for a single ping, we still observe a positive slope of the frequency response indicative of non-swimbladdered fish.

\begin{figure}
\includegraphics[width=16cm]{Fig_Sv_avg}
\caption{\label{Fig_Sv_avg} $Sv$ as a function of frequency ($m$) averaged over a range interval covering a fish school }
\end{figure}

>>>>

\newcommand{\code}[1]{\texttt{#1}}

@@

The theoretical foundation for signal processing comes from radar applications and has been documented for broadband echosounders \citep{stanton2008}. To tap the full potential of broadband echosounder data, further data processing methods are being developed \citep{lavery2017,bassett_broadband_2018}. Examples include adapting pulses to improve target separation as well as further parameters from the individual targets. Different transmission pulse forms can be envisioned as well as various methods for acoustic target classification. When translating equations into computer processing code, several choices must be made - some of these are well founded in signal processing literature whereas others are of a more practical and ad-hoc nature. The latter is typically missing in the literature, making it difficult to test the implementation of the signal processing methods in new echo sounders and post processing software. By presenting the design goals, implementation details, and recommended procedures and processing required to obtain quantitative broadband data, the authors hope to encourage and facilitate the realistic use of broadband signals in marine ecosystem acoustics.

@@

\subsection{Accompanying code}
The code accompanying this paper is written in the Python programming language and is available in the supplementary materials. All processing steps with respective figures in the paper can be reproduced by running the main scripts (\code{TSf.py} and \code{Svf.py}). Without loss of generality, we use the Simrad EK80 echosounder as an example since it is currently the most commonly used broadband system in the marine ecosystem acoustics field.

@@

\subsection{System overview}
A basic quantitative echosounder system consists of a transducer, a transceiver, and a computer program that controls the operation of the transceiver and records received signals. During transmission the program defines the signals that are created as electric signals in the transceiver, converted to acoustic signals by the transducer and transmitted into the water. The acoustic signals propagate through the water, are reflected or scattered by objects in the water, and propagate back to the transducer. During reception the transducer converts the received acoustic signals to electric signals, which are received, pre-amplified, filtered, digitized, processed in the transceiver, and then transferred to the controlling program for further data processing and storage (Fig. \ref{fi:ek_sys}). Many types of transmit signals are feasible - this paper considers only linear frequency modulated signals (also known as linear chirps).

@@

The controlling computer program generates a short-duration digital transmit signal (a ping), $\ytx(\samplesymt)$, where $\samplesymt$ is a sample index in the discrete time domain. Typical broadband pulses are linear upsweep pulses windowed by an envelope function. The generated signal $\ytxe(\timesym)$ is converted to an analogue electric signal and amplified by the transceiver to obtain an analogue signal $\ytxe(\timesym)$. The analogue and amplified signals are passed on to the transducer to generate the transmitted acoustic signal $\ytxa(\timesym)$ in the water, where $\timesym$ is the time for the (analog) signal. For a split-aperture echosounder system, there are typically three or four channels to allow estimation of the angle of arriving echoes, and the signal is typically transmitted with equal power across the channels.

In the example a linear sweep enveloped by a Hann function is implemented (Fig.~\ref{fi:ytx}), where the parameters are the initial frequency, the final frequency, the pulse duration, the sampling rate, and the proportion of the signal that is tapered in each end, respectively. A tapering of 0 and .5 indicates no tapering and tapering across the whole signal, respectively.
\begin{figure}
\includegraphics[width=16cm]{Fig_ytx}
\caption{\label{fi:ytx} Linear chirp pulses from 34 kHz to 45 kHz with a slope of 0.057 (blue) and 0.5 (orange), as implemented in the EK80.}
\end{figure}

@@

To remove noise and reduce the quantity of data, the sampled signal from each channel is filtered and decimated in multiple stages, $\stagesym$, using complex bandpass filters, $\hfl(\genidxsym,\stagesym)$, and decimation factors, $\decfac(\stagesym)$. The individual filter coefficients for each filter and decimation stage are indexed by $\genidxsym$. The output signal from each channel, $\channelsym$, from each filter and decimation stage, $\stagesym$, is then given by:
%
\begin{equation}
\label{eq:yrx}
\yrx(\samplesymt,\channelsym,\stagesym) = \left( \yrx(\samplesymt,\channelsym,\stagesym-1) * \hfl(\genidxsym,\stagesym) \right)_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\yrx(\samplesymt,\channelsym,0)$ is set to $\yrxorg(\samplesymt,\channelsym)$, being the signal before decimation, $*$ indicates convolution and $\nstages$ is the total number of filter stages. The output signal from the final filter and decimation stage, $\yrx(\samplesymt,\channelsym,\nstages)$, is shortened to $\yrx(\samplesymt,\channelsym)$ for convenience. For the output signal, $\yrx(\samplesymt,\channelsym)$, the decimated sampling rate, $\fsdec$, is given by:
%
\begin{equation}
\label{eq:fsdec}
\fsdec = \fs\prod_{\stagesym=1}^{\nstages} \frac{1}{\decfac(\stagesym)}.
\end{equation}
%
The characteristics of the bandpass filter and decimation factors are chosen with regard to the desired operating bandwidth, noise suppression levels, impulse response duration, and other common filter characteristics, with the aim of maintaining sufficient information in the data. In the example $N_v=2$. The frequency responses of the filters are shown in Figure~\ref{fi:fir} and the corresponding filter coefficients and decimation factors are given in the test data set.

@@

The original sample data $\yrxorg(\samplesymt,\channelsym)$ are not available in the EK80 data files. Instead, the filtered and decimated complex samples from each transducer channel $\yrx(\samplesymt,\channelsym)$ are stored in the data files. The data are recorded in computer data files for display and analysis by processing software. Additional information, such as that from position and motion sensors and system configuration data is also included in the files.

\subsection{Pulse compression}
To increase signal-to-noise ratio and resolution along the acoustic beam a matched filter may be applied to the raw data samples \citep{turin1960}. This technique is also known as pulse compression \citep{klauder1960}. One approach for a matched filter is to use a normalized version of the ideal transmit signal as the replicate signal, filtered and decimated using the same filters and decimation factors as applied in Eq. \ref{eq:yrx}. The normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, is given by:
%
\begin{equation}
\label{eq:ytxnorm}
\ytxnorm(\samplesymt) = \frac{\ytx(\samplesymt)}{\textrm{max}(\ytx(\samplesymt))}\end{equation}
%
where $\textrm{max}$ is the maximum value of $\ytx(\samplesymt)$. The filtered and decimated output signal, $\ytxnorm(\samplesymt,\stagesym)$, from each filter stage, $\stagesym$, using the normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, as the input signal, is given by:
%
\begin{equation}
\label{eq:FilterStagesTX}
\ytxnorm(\samplesymt,\stagesym) = \left[ \ytxnorm(\samplesymt,\stagesym-1) * \hfl(\genidxsym,\stagesym) \right]_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\ytxnorm(\samplesymt,0)$ is set to $\ytxnorm(\samplesymt)$ and $\downarrow$ indicates decimation by the factor $\decfac(\stagesym)$. The output signal from the final filter and decimation stage, $\ytxnorm(\samplesymt,\nstages)$, is used as the matched filter and is denoted as $\ymf(\samplesymt)$  (Fig.~\ref{fi:y_mf_n}).
%
\begin{figure}
\includegraphics[width=16cm]{Fig_y_mf_n}
\caption{\label{fi:y_mf_n} The absolute value of the filtered and decimated output signal, $\ymf(\samplesymt)$, that is used for the pulse compression.}
\end{figure}

@@

To perform pulse compression the received signal, $\yrx(\samplesymt,\channelsym)$, is convolved with a complex conjugated and time-reversed version of the matched filter signal with the matched filter signal, and here also normalized with the $l^2$-norm of the matched filter to maintain received signal power. The pulse compressed signal, $\ypc(\samplesymt,\channelsym)$, then becomes
\begin{equation}
\label{eq:PulseComp}
\ypc(\samplesymt,\channelsym) = \frac{\yrx(\samplesymt,\channelsym)*\ymf^*(-\samplesymt)}{||\ymf||^2_2},
\end{equation}
%
where $||\ymf||$ indicates the $l^2$-norm of $\ymf$, also known as the Euclidean norm. The received power samples are then used to estimate target strength and volume backscattering strength. For estimating received power samples, the mean signal, $\ypc(\samplesymt)$, over all transducer sectors, $\nchannels$, is used:
%
\begin{equation}
\label{eq:SumSig}
\ypc(\samplesymt) = \frac{1}{\nchannels} \sum_{\channelsym = 1}^{\nchannels} \ypc(\samplesymt,\channelsym).
\end{equation}
%
Compensation of echo strength for position in the acoustic beam requires an estimate of the echo arrival angle. This is obtained using the split-aperture method \citep{burdic1991}, which for broadband pules can be implemented with the angle values contained in the complex-valued $\ypc(\samplesymt)$ data, in combination with knowledge of transducer sector geometry. The principle is demonstrated with a transducer that is divided into four quadrants (Fig. \ref{fi:trd_quad}). In this example the summed signals from four halves (1+2, 2+3, 3+4, 4+1) are calculated as:
\begin{eqnarray}
\label{eq:SumHalves}
y_{\textrm{pc,fore}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,3)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,aft}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,2) \right),\\
y_{\textrm{pc,star}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,port}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,2)+\ypc(\samplesymt,3) \right),
\end{eqnarray}
%
where fore, aft, star(board), and port indicate the relevant transducer halves.
\begin{figure}
\includegraphics{Fig_trd_quad}
\caption{\label{fi:trd_quad}Transducer divided into four quadrants. The labels are directions often used when a transducer is mounted on a ship.}
\end{figure}
%

\subsection{Power and angle samples}
The transceiver measures voltage over a load, $\zrxe$, connected in series with the transducer impedance, $\ztde$. When calculating various acoustic properties, a system gain parameter will be used which assumes a matched receiver load. The total received power, $\prxe(\samplesymt)$, from all transducer sectors for a matched receiver load (Fig. \ref{fi:impedances}) is given by: 
\begin{equation}
\label{eq:prx}
\prxe(\samplesymt) = \nchannels\left( \frac{|\ypc(\samplesymt)|}{2\sqrt{2}} \right)^2 \left( \frac{|\zrxe+\ztde|}{\zrxe} \right)^2 \frac{1}{|\ztde|}.
\end{equation}
%
\begin{figure}
\includegraphics[width=16cm]{Fig_impedances}
\caption{\label{fi:impedances}Equivalent circuit diagram of transducer/transceiver with system impedances.}
\end{figure}

Forward/aft and port/starboard phase angles of target echoes are estimated by combining the transducer half signals thus: 
%
\begin{eqnarray}
\label{eq:phase1}
y_\along(\samplesymt) & = & y_{\textrm{pc,fore}}(\samplesymt) y_{\textrm{pc,aft}}^*(\samplesymt), \\
y_\athw(\samplesymt) & = & y_{\textrm{pc,star}}(\samplesymt) y_{\textrm{pc,port}}^*(\samplesymt),
\end{eqnarray}
%
where $y_\along(\samplesymt)$ is the electrical angle along the minor axis of the transducer (positive in the forward direction when ship-mounted) and $y_\athw(\samplesymt)$ the electrical angle along the major axis of the transducer (positive to starboard when ship-mounted), where complex signals are represented in the form $e^{j 2\pi \freqsym \timesym}$, where $j = \sqrt{-1}$. The physical echo arrival angles ($\along$ and $\athw$) are then given by:
%
\begin{eqnarray}
\label{eq:phase2}
\along(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\along(\samplesymt)), \Re(y_\along(\samplesymt) \right)}{\anglefalong}\right) \\
\athw(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\athw(\samplesymt)), \Re(y_\athw(\samplesymt) \right)}{\anglefathw}\right).
\end{eqnarray}
%
where $\anglefalong$ and $\anglefathw$ are constants that convert from phase angles to physical echo arrival angles and are derived from the transducer geometry and, $\fc$, the centre frequency of the chirp pulse \citep{ehrenberg1979}. The inverse sine is indicated by $\arcsin$,the four quadrant inverse tangent which returns values in the interval $[-\pi, \pi]$ inclusive is indicated by $\atan$, the real part of a complex number by $\Re$ and the imaginary part by $\Im$. As a mnemonic, the horizontal line in the symbol used for the forward/aft direction, $\along$, represents the pivot axis for the alongship angles and the near-vertical line in the $\athw$ symbol indicates the pivot axis for port/starboard angles.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_theta_phi}
\caption{\label{fi:theta_phi}Physical angles $\along$ and $\athw$.}
\end{figure}

@@

To illustrate each step of the process in obtaining $TS(m)$ we use a single ping from a data file where a 35 mm diameter tungsten carbide calibration sphere (WC35) was suspended approximately 6 m below a 120 kHz transducer (Fig.~\ref{Fig_TS_echogram}).

@@

Based on the point scattering strength samples and the phase angle samples, single targets can be detected, and range and bearing to the single targets can be estimated. This is typically achieved through a single echo detection algorithm (SED). Here we will assume that the samples from the pulse compressed data $\ypc(\samplesymt)$ originating from single target already have been identified, noting that the number of samples after the detected target may be higher than those those before the peak to include scattering processes that occur in actual targets (as opposed to ideal point targets). The alongship angle $\along_n$, athwartship angle $\athw_n$ and range $r_n$ at the \emph{peak} power $\prxe(\samplesymt)$ within the detected target are used as estimates for $\along_t$, $\athw_t$ and $r_t$, respectively (Fig. \ref{fi:TS}). A simple pseudo SED algorithm is implemented in the code for illustrative purposes.

From the autocorrelation function of the matched filter signal, $\ymfauto(\samplesymt)$, the equivalent number of samples around the peak are extracted to create the reduced autocorrelation signal of the matched filter signal, $\ymfautored(\samplesymt)$. Depending on the target scattering characteristics and the distance to any adjacent single targets, the number of samples around the peak echo level in $\ypctarget(\samplesymt)$ that contain the majority of the echo energy can be more or less than the total number of samples around the peak of $\ymfauto(\samplesymt)$. If the number of samples around the target is more than the total number of samples around the peak of $\ymfauto(\samplesymt)$ all samples around the peak of $\ymfauto(\samplesymt)$ are used. If the number of samples around the target is less than the total number of samples around the peak of $\ymfauto(\samplesymt)$, this lower number is used to create a reduced autocorrelation signal, $\ymfautored(\samplesymt)$.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_singleTarget}
\caption{\label{fi:SED} The $\prxe(\samplesymt)$ power (upper) and split beam angles ($\along_t$ and $\athw_t$) (middle) for the single target. The orange vertical line corresponds to the range $r_t$ for the single target. The $\ymfautored(\samplesymt)$ (lower) is the autocorrelation function of the transmit signal reduced to the length of the target signal and aligned with the peak power of the target.}
\end{figure}

@@

As a metallic sphere is a rather simple and near-ideal scatterer, we use $\sv$ from a school of non-swimbladdered fish (fig.~\ref{Fig_Sv_echogram}) collected with a 120 kHz centre frequency transducer to illustrate the processes to obtain $\sv(m)$.

@@

\begin{figure}
\includegraphics[width=16cm]{Fig_Sv_echogram}
\caption{\label{Fig_Sv_echogram} Sv as a function of frequency ($samplesymf$) and range ($samplesymt$) for a echogram raw data file. The red vertical line indicates the ping processed in this section.}
\end{figure}

@@

By selecting a set of centre samples $\timesym$, $\sv$ values can be presented as a function of range ($\samplesymt$) and frequency ($freqsym$) for each ping. The range for centre samples $\samplesymt$ could be chosen to half the window length or any other grid that the user prefer the data presented to be in. This can be useful when combining the $\sv(\freqsym)$ across a range of transducers. In our example we have simply chosen n to be on the same order of the range resolution (fig.~\ref{Fig_Sv_m_n}).

\begin{figure}
\includegraphics[width=16cm]{Fig_Sv_m_n}
\caption{\label{Fig_Sv_m_n} Sv as a function of frequency ($\samplesymf$) and range ($\samplesymt$) for a single ping}
\end{figure}

For acoustic abundance estimation and classification purposes it is common to integrate the $\sv$ over a range (15 to 34 m in the example, covering a school of non-swimbladdered fish, Fig.~\ref{Fig_Sv_echogram}). It is normal to average $\sv$ over several pings to obtain an unbiased estimate, but here just one ping is used for illustrative purposes (fig.~\ref{Fig_Sv_avg}). Even though this is for a single ping it is still possible to observe a positive slope of the frequency response that is indicative of non-swimbladdered fish.

\begin{figure}
\includegraphics[width=16cm]{Fig_Sv_avg}
\caption{\label{Fig_Sv_avg} $\sv$ as a function of frequency ($\samplesymf$) averaged over a range interval covering a fish school.}
\end{figure}
